Probabilistic Deep Learning as a Framework for Progressive Development of Behavior Models

### 2025-01-23 docs/Paper.tex
wip

\begin{center}
The purpose of this paper is to (1) show folks that they can use probabilistic deep learning to drive a virtuous cycle of progressive model development where they always have a useful model as they go along and (2) give them a guide as to exactly how to do it.
\end{center}

\begin{center}
This study explores using deep learning to build Individual-Based Models (IBMs). Using Chinook salmon (\textit{Oncorhynchus tshawytscha}) PSAT data as a case study, we introduce a methodology that leverages deep learning to transform the construction of IBMs into a systematic tool for investigating behaviors and their covariates. By employing a "log-odds model" to reduce dimensionality, this approach positions modeling as a central driver of discovery, enabling efficient hypothesis testing and iterative refinement.
\end{center}

@@

\section*{Applying Probabilistic Machine Learning to Behavior}

@@

Standard probabilistic deep learning networks are typically framed as a classification problem, using categorical cross-entropy as the loss function \citep{durr}. Each output neuron represents a potential choice, with the model predicting the probability of each choice being correct based on this loss formulation. For these choices, we provide the network with features encapsulating the relevant information. Training is then comprised of providing a series such decisions.

@@

To address this issue we could take advantage of the order invariance of choices in the traditional probabilistic problem framing. Specifically, the order in which choices are presented to the model should not matter. For instance, whether a particular choice appears in the first or the thirteenth position should have no impact on the model's operation. This property allows for data augmentation by reordering choices.

In essence, for each training example, $M!$ (factorial of the number of choices) augmented examples can be created.

The issue with this approach is that as your augmented data size grows to match the needs of the greater dimensionality, so too does the time required for training. So while the problem remains theoretically possible, the potential exponential uptick in time complexity poses a significant practical issue.

@@

We can implement this log-odds model using standard probabilistic deep learning techniques by replicating the "log-odds model" weights across all $M$ choices. The outputs are fed into a softmax layer with $M$ units, where the layer's weights are set to the identity matrix and biases are set to zero. Using categorical cross-entropy as the loss function ensures compatibility with standard probabilistic deep learning while enabling us to train the log-odds weights and significantly reduce the problem's dimensionality.

@@

The primary risk in using contrasts is introducing bias by disproportionately sampling certain combinations of choices. To mitigate this, we randomly sample pairs from each decision, ensure an equal number of contrasts per decision, and an equal number of decisions per individual. This preserves the balance across the training data and avoids skewing the model's predictions.

@@

In deep learning the specific layer sizes, depth, optimizer, learning rate, and other parameters best suited to a particular problem are not usually known from the start. Therefore it is important to do hyperparameter tuning in order to discover the best parameters for a particular problem. Practically this means training large numbers of models - a process that is usually very compute intensive. Even with just 3 parameters each of which can take on five results in 125 distinct models. Furthermore as with any search over a non-convex space (as backpropogation is) the initial seed can have impacts on the final solution. Therefore, training deep learning models requires training many models per problem.

This issue can be overcome by virtue of the fact that these models can all be trained independently and therefore in parallel. This is a perfect use case for cloud compute where one can spin up, for a short time and for a reasonable cost, a large number of machines and therefore do all of the hyperparameter tuning in parallel.

@@

In summary the application of probabilistic modeling proposed here follows three steps:

\begin{enumerate}
\item Discretize the behaviors into distinct decisions $d_i$ and choices $c_i,j$. 
\item Resample the choices using contrast sampling
\item Apply log-odds modeling to the contrasts, taking advantage of cloud compute to efficiently search the hyperparameter space
\item Use the trained model to make inferences on the original decisions
\end{enumerate}

The following example application demonstrates these steps in practice.

@@

This example will involve building three models of Chinook salmon (\textit{Oncorhynchus tshawytscha}) movement that build upon one another. These specific models were chosen to illustrate the capabilities of this technique as an exploratory data analysis (EDA) tool. The features are described in the following table:

@@

We will begin by describing the data and associated features, then move onto a description of the building of the models, and finally, in the results and discussion, illustrate how the models can be used to explore the data.

@@

Environmental data was derived from the Global Ocean Biogeochemistry Hindcast dataset (10.48670/moi-00019) and the Global Ocean Physics Reanalysis (10.48670/moi-00021) from the E.U. Copernicus Marine Service Information. Net primary production (mg/m3/day) and mixed layer thickness (m) were aggregated per Uber h3 resolution 4 cell in the Northern Pacific. \newline

Movement heading in radians and distance to the centroid of the choice cell were computed and then mixed layer thickness and net primary production were joined to the choices on cell and day.

Distance was normalized to a range of 0-1 by division by 100, while mixed layer thickness and net primary production were both log-scaled and then centered at zero.

@@

The first step in any other machine learning problem is splitting training and validation sets. Given the low number of individuals in the sample and the intention to demonstrate the technique's value as an EDA tool it was decided to not maintain a hold out test set.

@@

The next step in building a log-odds model is to decide on the formulation of our choices. In our case we decided to grid space by Uber h3 cells at resolution 4. Specifically, the cell containing each salmon location from our data was identified and then, assuming a maximum travel distance of 100km (centeroid to centeroid) all adjacent cells within the 100km were identified as choices (including the currently occupied cell). In general this represented $\sim 19$ choices per decision with the intention being to predict the probability of moving to any particular cell. Training data was derived by identifying the actual cell moved to.

@@

The next step in building a log-odds model is determining the specifics of the contrast sampling. For this example, after inspecting the distribution of number of choices per salmon and number of choices per decision, we decided on random sampling (with replacement) 200 decisions per individual and 19 choices per decision.

@@

For each of the three models trained architectures/hyperparameters for the internal log-odds component of the model were parametrized in the following ways:

@@

5 separate seeds were used for each combination. Models were trained in Keras using an Adam optimizer for 100 epochs. Training was done on AWS Batch using Fargate instances of 2 vcpu's and 4 GB of memory. By taking advantage of AWS Batch,  models could be all trained in parallel allowing for short (~1 hour) turn around times.

>>>>

@@

\section*{Guidance on Applying the Technique}

@@

Standard probabilistic deep learning networks are typically framed as a classification problem, using categorical cross-entropy as the loss function \citep{durr}. Using such a loss function amounts to minimizing the negative log likelihood across the examples in the training data. Each output neuron represents a potential choice, with the model predicting the probability of each choice being correct based on this loss formulation. For these choices, we provide the network with features encapsulating the relevant information. Training is then comprised of providing a series such decisions.

@@

To address this issue we could take advantage of the order invariance of choices in the traditional framing. In other words the order in which choices are presented to the model should not matter. For instance, whether a particular choice appears in the first or the thirteenth position should have no impact on the model's operation. This property allows for data augmentation by reordering choices. In essence, for each training example, $M!$ (factorial of the number of choices) augmented examples can be created.

The issue with this approach is that as your augmented data size grows to match the needs of the greater dimensionality, so too does the time required for training. So while the problem remains theoretically possible, the potential exponential uptick in time complexity poses a significant practical issue. \newline

\begin{figure}[h!] 
	\centering
  \includegraphics[height=55mm]{figures/log_odds.png}
  \caption{Log-Odds Model: Each choice is passed through a log odds model (left) whose output is then passed through a softmax layer whose weights are fixed in order to produce probabilities that can be passed to a categorical cross entropy loss function. By sharing weights between choices we effectively reduce the dimensionality of the problem. }
  \label{fig:log_odds}
\end{figure}

@@

We can implement this log-odds model using standard probabilistic deep learning techniques by replicating the "log-odds model" weights across all $M$ choices (Figure \ref{fig:log_odds}). The outputs are fed into a softmax layer with $M$ units, where the layer's weights are set to the identity matrix and biases are set to zero. Using categorical cross-entropy as the loss function ensures compatibility with standard probabilistic deep learning while enabling us to train the log-odds weights and significantly reduce the problem's dimensionality.

@@

The primary risk in using contrasts is introducing bias by disproportionately sampling certain combinations of choices. To mitigate this, we randomly sample pairs from each decision, ensure an equal number of contrasts per decision, and an equal number of decisions per individual. This preserves the balance across the training data and avoids skewing the model's learning.

@@

Finally, in deep learning the specific layer sizes, depth, optimizer, learning rate, and other parameters best suited to a particular problem are not usually known from the start. Therefore it is important to do hyperparameter tuning in order to discover the best parameters for a particular problem. Practically this means training large numbers of models - a process that is usually very compute intensive. Even with just 3 parameters 5 distinct values each results in 125 distinct models. Furthermore as with any search over a non-convex space, the initial seed can have impacts on the final solution. Therefore, training deep learning models requires training many models per problem.

This challenge can be overcome by virtue of the fact that these models can all be trained independently and therefore in parallel. This is a perfect use case for cloud compute where one can spin up, for a short time and a reasonable cost, a large number of machines and do all of the hyperparameter tuning in parallel. A primary benefit of the code at https://github.com/networkearth/mimic is to help do exactly this.

@@

In summary the steps in applying log-odds modeling are:

\begin{enumerate}
\item Discretize the behaviors into distinct decisions $d_i$ and choices $c_i,j$. 
\item Resample the choices using contrast sampling
\item Apply log-odds modeling to the contrasts, taking advantage of cloud compute to efficiently search the hyperparameter space
\item Use the trained model to make inferences on the original decisions (as opposed to the contrasts)
\end{enumerate}

@@

With this guide in hand we will proceed to demonstrate this technique by building three, very simply, models of Chinook salmon (\textit{Oncorhynchus tshawytscha}) movement. These specific models build on one another and were chosen to illustrate how one might use this technique as an exploratory data analysis (EDA) tool. The features for each model are as follows (the + indicating the model takes on the features of the model above it in addition to those listed in its row):

@@

We will begin by describing the data and associated features, then move onto the specifics of training the models (following our guide), and finally, in the results and discussion, illustrate how the models can be used to explore the data.

@@

\begin{figure}[h!] 
	\centering
  \includegraphics[height=55mm]{figures/mlt_2021_07_02.png}
  \caption{Mixed layer thickness (m) per h3 cell (resolution 4) on July 2, 2021.}
  \label{fig:mlt}
\end{figure}

Environmental data was derived from the Global Ocean Biogeochemistry Hindcast dataset (10.48670/moi-00019) and the Global Ocean Physics Reanalysis (10.48670/moi-00021) from the E.U. Copernicus Marine Service Information. Net primary production (mg/m3/day) and mixed layer thickness (m) were aggregated per Uber h3 resolution 4 cell in the Northern Pacific. As a reference for the resolution of the data see Figure \ref{fig:mlt}. \newline

Movement heading in radians and distance to the centroid of the choice cell were computed. Mixed layer thickness and net primary production were joined to the choices on h3 cell index and day.

Distance was normalized to a range of 0-1 through division by 100, while mixed layer thickness and net primary production were both log-scaled and then centered at zero.

@@

The first step was to split into training and validation sets. Given the low number of individuals in the sample and the intention to demonstrate the technique's value as an EDA tool it was decided to not maintain a hold out test set.

@@

The next step in building a log-odds model is to decide on the formulation of our choices. In our case we decided to grid space by Uber h3 cells at resolution 4. Specifically, the cell containing each salmon location from our data was identified and then, assuming a maximum travel distance of 100km (centroid to centeroid) all adjacent cells within the 100km were identified as choices (including the currently occupied cell). In general this represented $\sim 19$ choices per decision. As such our model ends up predicting the probability, given the data, of moving to any one of those cells. Training data was derived by identifying the actual cell moved to.

@@

Next we needed to determine the specifics of the contrast sampling. For this example, after inspecting the distribution of number of choices per salmon and number of choices per decision, we decided on random sampling (with replacement) 200 decisions per individual and 19 choices per decision.

@@

For each of the three models trained, the hyperparameters for the internal log-odds component of the model were parametrized in the following way:

@@

We proceeded by grid search and used 5 separate seeds for each combination. Models were trained in Keras using an Adam optimizer for 100 epochs. Training was done on AWS Batch using Fargate instances of 2 vcpu's and 4 GB of memory. By taking advantage of AWS Batch,  models could be all trained in parallel allowing for short (~1 hour) turn around times.
